Assumptions of Physics blueprint: information granularity as a common foundation for geometry, measures and probability

### 2021-04-23 Overviews/Informational granularity/InformationGranularity.tex
InfGran: review with Mark

\begin{prop}
Other properties satisfied by fineness:
\begin{enumerate}
    \item if $\stmt[s]_1 \ncomp \stmt[s]_2$ and $\stmt[v]_1 \ncomp \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ and $\stmt_2 \finer \stmt[v]_2$ imply  $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
    \item (monotony as in Villegas) if $\stmt[v]_1 \ncomp \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ and $\stmt_2 \finer \stmt[v]_2$ imply  $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
    \item if $\stmt_1 \ncomp \stmt_2$, $\stmt[v]_1 \ncomp \stmt[v]_2$ and $\stmt_2 \eqgran \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ if and only if $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
\end{enumerate}
\end{prop}

@@

@@

@@

\begin{conj}
	The relation $\lequ$ is reflexive and transitive and therefore a preorder.
\end{conj}

Fortunately, the current definitions of $\lequ$ and $\equ$ behave well together:

@@

@@

@@

\begin{defn}[Monotone continuity]
	Given a domain $\tdomain$ and a unit $\stmt[u] \in\tdomain$, we say that $\lequ$ satisfies \textbf{monotone continuity} if the following is true: given a sequence $\{\stmt_i\}_{i=1}^{\infty}$ such that $\stmt_i \narrower \stmt_j$ for all $ i \leq j$ and $\stmt = \bigOR_i\stmt_i$ the disjunction, for every $\stmt[t] \in \tdomain$ such that $\stmt_i\leq_{\stmt[u]} \stmt[t]$ for all $i$ we have $\stmt\leq_{\stmt[u]} \stmt[t]$.
\end{defn}

@@

We therefore want to make the following distinction. TODO: define countable vs uncountable

\begin{defn}
	We say that $\stmt$ is a \textbf{uncountable} infinitesimal with respect to $\stmt[u]$ if all there exists a sequences $\{ \stmt_i \}_{i=1}^\infty$ such that:
	\begin{enumerate}
		\item $\stmt \finer \stmt_i$ for all $i$
		\item $\stmt_i \ncomp \stmt_j$ for all $i \neq j$
		\item $\bigOR\limits_{i=1}^n \stmt_i \finer \stmt[u]$ 
	\end{enumerate}
	In words, $\stmt$ is a countable infinitesimal with respect to $\stmt[u]$ if we can find for any number, there is a disjunction of that many broader mutually incompatible statements $\stmt_i$ is finer than $\stmt[u]$.
\end{defn}

TODO: desiderata that if there are no countable infinitesimal, then we have monotone continuity

We wish to understand the following. This is monotone continuity up to infinitesimals, and so with the right definitions, it should translate to monotone continuity for $\lequ$.

Let $\stmt[u]$ be a statement which does not admit countable infinitesimals, and let $\{\stmt_i:i\in\mathbb{N}\}$ be a sequence of statements monotone increasing in narrowness (i.e. $\stmt_i\narrower\stmt_j$ for all $i\leq j$). Write $\stmt=\OR_i\stmt_i$ for the disjunction and assume $\stmt$ is finitely comparable to $\stmt[u]$. If $\stmt[t]\in\tdomain$ such that $\stmt[t]$ is finitely comparable to $\stmt[u]$, and $\stmt_i\finer\stmt[t]$ for all $i$, then $\stmt\finer\stmt[t]\OR\stmt[w]$ for some infinitesimal (with respect to $\stmt[u]$) $\stmt[w]$.

@@

\subsection{Some thoughts on Example in 3.3}

The construction above hints at some fundamental concepts in measure theory which will be important for us to consider.

Notice that the two sets $V$ and $W$ are both of infinite Lebesgue measure (area). With respect to Lebesgue measure, monotone continuity is immediate since if $\bigcup\limits_0^n W\finer V$ for all $n$ then $\lim\limits_{n\to\infty}\mu(\bigcup\limits_0^n W) = \mu(W)\finer V$. Even the above example does not violate this, since there, both would be ``equivalent" in fineness, namely they are infinitely coarse.

Our goal is to have a measure which is able to distinguish sets which may be of infinite Lebesgue measure as well. This will be very difficult to fully capture. Consider the following procedure. Starting with the vertical strip $V$, slide all the squares at even $y$-values to the right by one, and then allow the squares to collapse down to rest on the square below. Now, by simply rearranging regions, we have (almost) a cover of the wider strip $W$. The only missing points are the line $x=1, y>0$. If instead we used $(0,1]$ and $(0,2]$ for the strips, this procedure would cover them exactly. One could continue and cover the entire plane with similar translations.

Here is one approach to distinguish $V$ and $W$, and similar methods will enable one to distinguish many more infinite-measure subsets. Let $S_n$ be the $2n\times 2n$ square centered the origin with vertices at $(\pm n, \pm n)$. Then for all $n$, $V\cap S_n$ has area $n$, and $W\cap S_n$ has area $2n$. This holds for all $n$ and so we might use something like this to think about how we would distinguish infinite areas. Notice that this is in some sense writing $V$ and $W$ as countable unions of nested sets in a standardized way (e.g. $V = \bigcup\limits_n V\cap S_n$) and then comparing the finite-area subsets. One could choose shapes other than $S_n$ for this purpose. Another way to distinguish them would be to look at the limit $\lim\limits_{n\to\infty}\text{Area}(V\cap S_n)/\text{Area}(S_n)$; in this case they are both 0 but for a different choice of $S_n$ one could find an asymptotic difference. 
The point of the above discussion is to try to gain a better idea of which (infinite/unbounded) sets we want to be related by $\finer$. It is not clear even for ``nice" regions of $\mathbb{R}^2$, let alone at the full generality we are seeking.

Regardless, the goal is not just to distinguish infinite-Lebesgue-measure sets. We need to understand when monotone continuity is relevant and what happens without it.

@@

Consider the subset $[0,\infty)\times[0,\infty)\subseteq\mathbb{R}^2$. Let $U_{i,j} = [i, i+1)\times[j,j+1)$ be the unit square with bottom-left corner $(i,j)$. Let $V$ be the vertical (infinite) strip $V = [0,1)\times[0,\infty)$ and $W = [0,2)\times[0,\infty)$. Define the sequence $W_i = \{ U_{0,0}, U_{1,0}, U_{0,1}, U_{1,1}, U_{0,2}, U_{1,2}, U_{0,3}, U_{1,3}, ... \}$ building upwards row by row. We have $\bigcup\limits_0^\infty W_i = W$. This means $\bigcup\limits_{i=0}^n W_i\finer V$ for all finite $n$ but $\bigcup\limits_{i=0}^\infty W_i = W \coarser V$.

TODO: show that this is a problem for $\lequ$

@@

TODO: Make this an example of how measure zero is not equigranularity

@@

@@

\begin{defn}
Let $(S,\leq)$ be a poset. A \textbf{chain} is a subset $T\subseteq S$ which is totally ordered (i.e. every pair of elements in $T$ is comparable in $\leq$). A chain is said to be \textbf{maximal} if it is not a proper subset of any other chain, i.e. there are no ``gaps" in the chain. 
\end{defn}
Observe that in a totally ordered set, there is only one maximal chain, namely the entire set.

@@

The issue is that the granularity of a countable sequence with each term a finite size but an infinite limit ``jumps'' at infinity. For example, take the sequence $\{1\}$, $\{1, 3\}$, $\{1, 3, 5\}$, ... , \{ all odd natural numbers\}. Compare to $\{1\}$, $\{1, 2\}$, $\{1, 2, 3\}$, ... , \{ all natural numbers\}. They are both maximal chains. They both go through all sizes of finite sets, but they reach infinity at two different levels. Effectively, the above property is only true over finite sets.

The following consequences of the MCP are interesting, but only use much weaker statements about the fineness classes of possibilities.

@@

TODO: add remark that MCP -> finite domain

\begin{remark}
The above observations and propositions show that MCP and comparability of possibilities imply equigranular possibilities, but equigrangular possibilities and totality do \emph{not} imply MCP.

Further, MCP alone implies neither totality nor equigranular possibilities. Indeed, consider a theoretical domain with only two possibilities and a fineness ordering which has two unequal successors to the fineness of $\impossibility$. This satisfies the MCP but fails the other conditions. 
\end{remark}

>>>>

\begin{prop}\label{fineness_monotony}
Other properties satisfied by fineness:
\begin{enumerate}
    \item if $\stmt[s]_1 \ncomp \stmt[s]_2$ and $\stmt[v]_1 \ncomp \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ and $\stmt_2 \finer \stmt[v]_2$ imply  $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
    \item (monotony as in Villegas) if $\stmt[v]_1 \ncomp \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ and $\stmt_2 \finer \stmt[v]_2$ imply  $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
    \item if $\stmt_1 \ncomp \stmt_2$, $\stmt[v]_1 \ncomp \stmt[v]_2$ and $\stmt_2 \eqgran \stmt[v]_2$, then $\stmt_1 \finer \stmt[v]_1$ if and only if $\stmt_1 \OR \stmt_2 \finer \stmt[v]_1 \OR \stmt[v]_2$.
\end{enumerate}
\end{prop}

@@

\begin{defn}
	Let $\stmt_1, \stmt_2 \in \tdomain$ be two statements such that $\stmt_1 \finer \stmt_2$. We say that they are \textbf{finitely comparable} to each other if we can find a finite set $\{ \stmt[t]_i \}_{i=1}^n$ such that $\stmt[t]_i \finer \stmt_1$ for all $i \leq n$ and $\stmt_2 \finer \bigOR_{i=1}^n \stmt[t]_i$.
\end{defn}

\begin{prop}
	If two statements are finitely comparable then they cannot be infinitesimals with respect to each other.
\end{prop}
\begin{proof}
	TODO
\end{proof}

@@

The following property is crucial for the theory of qualitative probability, but fortunately it is immediate from either definition.

\begin{prop}
For all statements $\stmt$ and $\stmt[u]$, we have $\impossibility\lequ\stmt\lequ\certainty$.
\end{prop}
\begin{proof}
	TODO
\end{proof}

@@

\begin{desid}
	The relation $\lequ$ is reflexive and transitive and therefore a preorder. 
\end{desid}

Fortunately, at least the first definitions of $\lequ$ and $\equ$ behave well together:

@@

For the second definition of $\lequ$, transitivity of $\equ$ is needed which we have not proven.

@@

Eventually, we wish to use $\lequ$ as a total pre-order in order to use it as a qualitative probability. We will need an interesting-enough subset of $\tdomain$ to capture the desired notions while still having $\lequ$ as a total order.

@@

\begin{defn}[Monotone continuity]\label{monotone_continuity}
	Given a domain $\tdomain$ and a unit $\stmt[u] \in\tdomain$, we say that $\lequ$ satisfies \textbf{monotone continuity} if the following is true: given a sequence $\{\stmt_i\}_{i=1}^{\infty}$ such that $\stmt_i \narrower \stmt_j$ for all $ i \leq j$ and $\stmt = \bigOR_i\stmt_i$ the disjunction, for every $\stmt[t] \in \tdomain$ such that $\stmt_i\leq_{\stmt[u]} \stmt[t]$ for all $i$ we have $\stmt\leq_{\stmt[u]} \stmt[t]$.
\end{defn}

@@

We therefore want to make the following distinction.

\begin{defn}
We say that $\stmt$ is a \textbf{countable infinitesimal} with respect to $\stmt[u]$ if it satisfies the following conditions:
\begin{enumerate}
    \item $\stmt$ is an infinitesimal with respect to $\stmt[u]$
    \item there exists a (countable) sequence $\{\stmt_i\}_{i=1}^{\infty}$ such that $\stmt_i\finer\stmt$ for all $i$, and $\bigOR\limits_{i=1}^{\infty}\stmt_i$ is comparable but not infinitesimal with respect to $\stmt[u]$
\end{enumerate}
An infinitesimal which is not a countable infinitesimal will be called an \textbf{uncountable infinitesimal}. 
\end{defn}

An example of a space with countable infinitesimals is the natural numbers $\mathbb{N}$. If the possibilities are natural numbers themselves and they are equigranular, for any infinite subset $S$ of $\mathbb{N}$, each natural number is infinitesimal with respect to $S$. However, a countable sequence of natural numbers may not be infinitesimal with respect to $S$ (and indeed may surpass $S$ in coarseness). Another example of countable infinitesimals is any subset of $\mathbb{R}$ of finite positive Lebesgue measure.

An example of a space with no countable infinitesimals with respect to the full space is an interval of finite length in $\mathbb{R}$ (any positive-Lebesgue measure subset will work). To see why, note that all infinitesimals are measure zero, and any countable union of measure zero sets is still measure zero.

It seems possible that the only obstruction to monotone continuity is countable infinitesimals, since they allow limits of equigranular sequences to be of different granularity classes.

\begin{desid}
    If $\stmt[u]$ is such that there are no countable infinitesimals, then $\lequ$ has monotone continuity. 
\end{desid}

Before achieving the above, it may be possible to obtain a related result about fineness and monotone continuity, which could translate to monotone continuity for $\lequ$.

\begin{desid}
Let $\stmt[u]$ be a statement which does not admit countable infinitesimals, and let $\{\stmt_i\}_{i=1}^{\infty}$ be a sequence of statements monotone increasing in narrowness (i.e. $\stmt_i\narrower\stmt_j$ for all $i\leq j$). Write $\stmt=\OR_i\stmt_i$ for the disjunction and assume $\stmt$ is finitely comparable to $\stmt[u]$. If $\stmt[t]\in\tdomain$ such that $\stmt[t]$ is finitely comparable to $\stmt[u]$, and $\stmt_i\finer\stmt[t]$ for all $i$, then $\stmt\finer\stmt[t]\OR\stmt[w]$ for some infinitesimal (with respect to $\stmt[u]$) $\stmt[w]$. 
\end{desid}

Tentative layout:
- summarize the result in terms of input requirements (stated with our notation) and output
- discussion of how far we are from getting those requirements
- sketch of the main steps of the proof (the key propositions and the approach of the proof of each in our notation)
- discussion on whether it is feasable and how to relax the proof for finite additivity

The paper \cite{villegas} by Villegas gives what seems to be an extremely relevant overview of qualitative probability, including many of the axioms which we seek in our structures.

In this part, we will assume that $\tdomain_{\stmt[u]}$ is a subset of the theoretical domain on which $\lequ$ is total, $\stmt[u]\in \tdomain_{\stmt[u]}$ is a distinguished element (we'll refer to it as the \emph{unit} when needed). We will use the word ``equiprobable" for $=_{\stmt[u]}$ since it means something different from equigranular (since $\leq_{\stmt[u]}$ is a different ordering from fineness, even though it will be derived from fineness).

The goal is to use $\lequ$ as a qualitative probability. The first basic requirement is that $\leq_{\stmt[u]}$ is a total pre-order with maximal and minimal elements. These were contained in several desiderata earlier.

Below is the version of the Axiom of Monotony for $\lequ$; earlier in Proposition \ref{fineness_monotony} it was shown that fineness satisfies this, and Definition \ref{smalleru} would allow this to descend to $\lequ$ as well.

The following are consequences of $\lequ$ being a total pre-order with the Axiom of Monotony. Let $\stmt_1,\stmt_2,\stmt_3,\stmt_4\in \tdomain_{\stmt[u]}$.
\begin{itemize}
    \item If $\stmt_1\leq_{\stmt[u]} \stmt_2$ and $\stmt_2<_{\stmt[u]} \stmt_3$, then $\stmt_1<_{\stmt[u]}\stmt_3$. 
    \item If $\stmt_1\narrower \stmt_2$ then $\stmt_1\leq_{\stmt[u]} \stmt_2$
    \item If $\stmt_1\leq_{\stmt[u]} \stmt_2$, $\stmt_3\leq_{\stmt[u]} \stmt_4$, $\stmt_2\leq_{\stmt[u]} \stmt_4$, and $v\leq_{\stmt[u]} \stmt_1$, then $\stmt_2 \AND \NOT \stmt_1 \leq_{\stmt[u]} \stmt_4 \AND \NOT \stmt_3$
    \item If $\stmt_1\leq_{\stmt[u]} \stmt_2$ then $\NOT \stmt_2 \leq_{\stmt[u]}\NOT \stmt_1$
\end{itemize}

Beyond the above, we will also utilize Monotone Continuity (Definition \ref{monotone_continuity}) in much of what follows.

Another important notion for qualitative probability is the idea of atoms.

\begin{defn}
An \textbf{atom} is a statement $\stmt\in S$ such that $\impossibility<_{\stmt[u]}\stmt$ and there does not exist any statement $\stmt_2\in S$ such that $\impossibility<_{\stmt[u]} \stmt_2<_{\stmt[u]} \stmt$. 
\end{defn}
In our framework, if possibilities are equigranular and not equivalent at $\stmt[u]$ scale to $\impossibility$, then they will be atoms. The presence or lack of atoms has several important consequences. In particular, Theorem 5 in \cite{villegas} \S3 includes the fact that $\tdomain_{\stmt[u]}$ is atomless if and only if every statement can be partitioned into two equiprobable statements. Theorem 3 in \S4 states that if $\tdomain_{\stmt[u]}$ is atomless, then there exists a unique probability measure compatible with $\leq_{\stmt[u]}$, and it is countably additive.

A major goal of this work is to discover the right axioms and definitions that will lead to a notion of $\lequ$ which has the properties necessary for the existence (and uniqueness) of a compatible probability measure on $\tdomain_{\stmt[u]}$. The above axioms discussed are enough - several theorems from \cite{villegas} give the desired result (in particular, see Theorem 3 in Section 4 and Theorem 5(iv) in Section 3, as well as their proofs). However, we have not as of yet been able to prove that $\lequ$ has the right properties. It is also possible that our framework will allow us to make use of different sets of axioms that lead to a similar conclusion. The survey \cite{fishburnsurvey} contains some examples of different axioms which may be very helpful.

\subsection{Old notes}

@@

% END OF NOTES

@@

Consider the subset $[0,\infty)\times[0,\infty)\subseteq\mathbb{R}^2$. Let $U_{i,j} = [i, i+1)\times[j,j+1)$ be the unit square with bottom-left corner $(i,j)$. Let $V$ be the vertical (infinite) strip $V = [0,1)\times[0,\infty)$ and $W = [0,2)\times[0,\infty)$. Define the sequence $W_i = \{ U_{0,0}, U_{1,0}, U_{0,1}, U_{1,1}, U_{0,2}, U_{1,2}, U_{0,3}, U_{1,3}, ... \}$ building upwards row by row. We have $\bigcup\limits_0^\infty W_i = W$. This means $\bigcup\limits_{i=0}^n W_i\sfiner V$ for all finite $n$ but $\bigcup\limits_{i=0}^\infty W_i = W \scoarser V$. This is a violation of monotone continuity, including the definition given up to infinitesimals (since $\bigcup\limits_{i=0}^{\infty}W_i$ is more than an infinitesimal greater than $V$).

By either definition of $\lequ$, this would give us an increasing sequence $W_n' := \bigcup\limits_{i=0}^n$, all of whose terms are (strictly) bounded by $V$, but whose limit (strictly) exceeds $V$ (in $\lequ$ terms), and so this violates monotone continuity of $\lequ$.

@@

TODO(GC): Make this an example of how measure zero is not equigranularity

@@

Note that even two countable subsets of possibilities need not be equigranular. For example, if $\{\stmt[x]_i\}_{i=0}^{\infty}$ is a set of countably many (equigranular) possibilities, and $\stmt = \bigOR\limits_{i=0}^{\infty}\stmt[x]_i$, then $\stmt' = \bigOR\limits_{i=0}^{\infty}\stmt[x]_{2i}$ is strictly finer.

Note: I didn't quite recall what we were thinking with this TODO. But if we are wanting to show that two measure zero sets can be non-equigranular, we can go much more generally and not rely on a measure with the above example of countable sets.

@@

The problem is that infinite sets disrupt the relationship between narrowness and fineness, such that a maximal chain in narrowness is not necessarily a maximal chain in fineness. For example, if we have a countable set of possibilities (e.g. the natural numbers), a maximal chain in narrowness can start from the empty set, start by adding possibilities one be one until we end with the full set. The set with all possibilities except one, for example, cannot be added to the chain, but still represents a level of description that is in between all the finite sets and the full infinite set. The only case where we can construct maximal chains is precisely when the possibilities are finite, and is therefore not interesting.

\begin{defn}
Let $(S,\leq)$ be a poset. A \textbf{chain} is a subset $T\subseteq S$ which is totally ordered (i.e. every pair of elements in $T$ is comparable in $\leq$). A chain is said to be \textbf{saturated} if there are no ``gaps" in the chain. That is, there is no $s \in S$ such that $s \notin T$, $t_1 \leq s \leq t_2$ for some $t_1, t_2 \in T$ and $T \cup \{s\}$ is a chain. 
\end{defn}

TODO(GC): everything needs to be changed to saturated

@@

The following made the idea look promising.

@@

\begin{prop}
Let $\tdomain$ be a theoretical domain with the maximal chains property. Then they must have finitely many possibilities.
\end{prop}
\begin{proof}
Let $\tdomain$ be a theoretical domain with the maximal chains property and such that the set $X$ of possibilities is infinite. Let $X' = \{x_i\}_{i=0}^{\infty} \subseteq X$ be a countable subset. Let $s_j = \bigOR_{j=0}^{i} x_i$ be the disjunction of the first $j$ possibilities. Let $\stmt = \bigOR_{i=0}^{\infty} x_i$ be the disjuction of all the possibilities in the sequence. The set $C = \{ \impossibility, \stmt \} \cup \{\stmt_j\}_{j=0}^{\infty}$ is a chain. In fact $\impossibility \snarrower \stmt_j \snarrower \stmt$ for all $j$ and $\stmt_i \narrower \stmt_j$ if $i \leq j$.

To show that $C$ is maximal, take a statement $\hat{\stmt} \narrower \stmt$ such that $\hat{\stmt} \notin C$. Let $A \subseteq X'$ be the set of possibilities compatible with $\hat{\stmt}$. If $A$ is infinite, since $\hat{\stmt} \nequiv \stmt$, there exists an $x_i \in X'$ such that $x_i \notin A$. Therefore $\stmt_{i} \nnarrower \hat{\stmt}$ since the first is compatible with a possibility that is not compatible with the second, and $\hat{\stmt} \nnarrower \stmt_{i}$ since $\hat{\stmt}$ is compatible with infinitely many statements and $\stmt_{i}$. If $A$ is finite, let the $x_i \in A$ for which $i$ is highest. Note that $A$ cannot contain all previous $x_j$ in the sequence, since $\hat{\stmt} \notin C$. Let $x_j$ the possibility with the highest $j \leq i$ that is not in $A$. We have $\hat{\stmt} \nnarrower \stmt_{j}$ and $\stmt_{i} \nnarrower \hat{\stmt}$ since each is compatible with a possibility incompatible with the other. Since we cannot add other statement, $C$ is maximal.

Now we show that, even though $C$ is maximal, it does not go through all levels of granularity. Let $x \in X'$ and consider $\stmt \AND \NOT x$. This is finer than $\stmt$ and is coarser than all other statements statements in $C$ since they are all compatible with finitely many possibilities. Therefore MCP does not hold. Therefore, if MCP holds, the domain cannot have infinitely many possibilities.
\end{proof}
